Thermodynamic response functions in a cell fluid model

### 2024-09-10 paper_text.tex
change notation for isothermal compressibility from K_T to \kappa_T

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	K_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T, N}.
\end{equation}
Let us perform some transformations to rewrite $K_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}:
\begin{eqnarray*}
	K_T	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,N}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}}.
\end{eqnarray*}
We have omitted the condition of constant $N$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $K_T$, one gets
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	K_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
\end{equation}

Let us introduce the reduced isothermal compressibility
\begin{equation}
	K^*_T \equiv \frac{\varepsilon K_T}{v}.
\end{equation}
The quantity $K^*_T$ is dimensionless and is of order unity, except at the critical point itself, where it is divergent. It is expressed in terms of the reduced quantities $P^*$ and $\rho^*$ as follows
\begin{eqnarray}
	K^*_T & = & \frac{\varepsilon}{{\rho^*}^2} \left(\frac{\partial \rho^*}{\partial \mu}\right)_T
	\nonumber\\
	& = & \frac{\varepsilon^2}{{\rho^*}^2} \left(\frac{\partial^2 P^*}{\partial \mu^2}\right)_T.
\end{eqnarray}
Now, either Eq.~\eqref{eq:eosPTM_reduced} or Eq.~\eqref{eq:density} can be used to explicitly calculate $K^*_T$, with the result expressed as a function of temperature and chemical potential.

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then the most suitable expressions for the isothermal compressibility are
\begin{equation}
	K_T = \frac{1}{\rho^*} \left(\frac{\partial P}{\partial \rho^*}\right)^{-1}_T
\end{equation}
and 
\begin{equation}
	K^*_T = \frac{1}{\rho^*} \left(\frac{\partial P^*}{\partial \rho^*}\right)^{-1}_T.
\end{equation}

Figure~\ref{fig2a} illustrates the dependency of the compressibility $K^*_T$ on the density $\rho^*$ for various values of temperature above the critical one. The dependency of $K^*_T$ on the effective chemical potential $M$ is displayed in Fig~\ref{fig2b}.

\begin{figure}[htbp]
	%\centering 
	\includegraphics[width=0.5\textwidth]{f2a.pdf}
	%\includegraphics[width=\column]{f2a}
	%\vskip-3mm
	\caption{The reduced isothermal compressibility $K^*_T$ as a function of the density $\rho^*$ at different values of reduced temperature $\tau > 0$ ($T > T_c$). 
	}
	\label{fig2a}
\end{figure}

\begin{figure}[h!]
	\centering \includegraphics[width=0.45\textwidth]{f2b.pdf}
	\includegraphics[width=0.475\textwidth]{f2c.pdf}
	\vskip-3mm\caption{The reduced isothermal compressibility $K^*_T$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
	}\label{fig2b}
\end{figure}

@@

\subsection{Relation between response functions}
The following thermodynamic identity holds between the calculated response functions
\begin{equation}
	\label{eq:identity}
	\frac{\alpha_P}{K_T \beta_V} = \frac{\alpha^*_P}{K^*_T \beta^*_V} = 1,
\end{equation}
which is easily derived from the cyclic relation between $P$, $V$ and $T$. In our case, this equality is exactly reproduced by substituting explicit expressions for $K_T$, $\alpha_P$, and $\beta_V$ obtained from the equation of state~\eqref{eq:eosMT} (or~\eqref{eq:eosNT}).

>>>>

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	\kappa_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T, N}.
\end{equation}
Let us perform some transformations to rewrite $\kappa_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}:
\begin{eqnarray*}
	\kappa_T	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,N}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}}.
\end{eqnarray*}
We have omitted the condition of constant $N$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $\kappa_T$, one gets
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
\end{equation}

Let us introduce the reduced isothermal compressibility
\begin{equation}
	\kappa^*_T \equiv \frac{\varepsilon K_T}{v}.
\end{equation}
The quantity $\kappa^*_T$ is dimensionless and is of order unity, except at the critical point itself, where it is divergent. It is expressed in terms of the reduced quantities $P^*$ and $\rho^*$ as follows
\begin{eqnarray}
	\kappa^*_T & = & \frac{\varepsilon}{{\rho^*}^2} \left(\frac{\partial \rho^*}{\partial \mu}\right)_T
	\nonumber\\
	& = & \frac{\varepsilon^2}{{\rho^*}^2} \left(\frac{\partial^2 P^*}{\partial \mu^2}\right)_T.
\end{eqnarray}
Now, either Eq.~\eqref{eq:eosPTM_reduced} or Eq.~\eqref{eq:density} can be used to explicitly calculate $\kappa^*_T$, with the result expressed as a function of temperature and chemical potential.

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then the most suitable expressions for the isothermal compressibility are
\begin{equation}
	\kappa_T = \frac{1}{\rho^*} \left(\frac{\partial P}{\partial \rho^*}\right)^{-1}_T
\end{equation}
and 
\begin{equation}
	\kappa^*_T = \frac{1}{\rho^*} \left(\frac{\partial P^*}{\partial \rho^*}\right)^{-1}_T.
\end{equation}

Figure~\ref{fig2a} illustrates the dependency of the compressibility $\kappa^*_T$ on the density $\rho^*$ for various values of temperature above the critical one. The dependency of $\kappa^*_T$ on the effective chemical potential $M$ is displayed in Fig~\ref{fig2b}.

\begin{figure}[htbp]
	%\centering 
	\includegraphics[width=0.5\textwidth]{f2a.pdf}
	%\includegraphics[width=\column]{f2a}
	%\vskip-3mm
	\caption{The reduced isothermal compressibility $\kappa^*_T$ as a function of the density $\rho^*$ at different values of reduced temperature $\tau > 0$ ($T > T_c$). 
	}
	\label{fig2a}
\end{figure}

\begin{figure}[h!]
	\centering \includegraphics[width=0.45\textwidth]{f2b.pdf}
	\includegraphics[width=0.475\textwidth]{f2c.pdf}
	\vskip-3mm\caption{The reduced isothermal compressibility $\kappa^*_T$ as a function of the effective chemical potential $M$ for different temperatures $\tau = (T - T_c)/T_c$ at $T > T_c$. The two figures differ in the scale of $M$. The left-hand-side figure covers a wider range of $M$, while the right-hand-side one focuses on a range of $M$ around its critical value $0$.
	}\label{fig2b}
\end{figure}

@@

\subsection{Relation between response functions}
The following thermodynamic identity holds between the calculated response functions
\begin{equation}
	\label{eq:identity}
	\frac{\alpha_P}{\kappa_T \beta_V} = \frac{\alpha^*_P}{\kappa^*_T \beta^*_V} = 1,
\end{equation}
which is easily derived from the cyclic relation between $P$, $V$ and $T$. In our case, this equality is exactly reproduced by substituting explicit expressions for $\kappa_T$, $\alpha_P$, and $\beta_V$ obtained from the equation of state~\eqref{eq:eosMT} (or~\eqref{eq:eosNT}).
